Mendelian randomization for causal inference in Genome-wide association studies

### 2022-02-14 mr-review.Rmd
added some biotech refs

Genome-wide association studies (GWAS) present unique opportunities for 
discovery of causal relationships via MR. A GWAS
interrogates millions of single nucleotide polymorphisms (SNPs) to infer which affect the trait [@the_wellcome_trust_case_control_consortium_genome-wide_2007;@klein_complement_2005].
In the nearly 17 years since publication of early studies, researchers have reported 
tens of thousands of novel SNP-trait associations from GWAS [@klein_complement_2005;@welter_nhgri_2014;@visscher_10_2017;@visscher_five_2012].

@@

Recently, multiple teams of investigators have analyzed time-to-event traits in MR studies [@burgess_commentary_2015;@arvanitis_linear_2021;@tikkanen_body_2019;@he_causal_2018;@guo_body_2017]. 
@guo_body_2017, for example, studied post-diagnosis survival time in women with breast cancer. They 
treated survival as a time-to-event outcome, which enabled them to model it with a proportional 
hazards model [@cox_regression_1972] in two-stage MR with individual-level data. Similarly, 
@he_causal_2018, in a study of cardiovascular disease, applied proportional hazards regression when 
examining time to stroke as an outcome variable. We are unaware of MR studies that treat 
the exposure variable as a time-to-event, but it's conceivable that future research will consider 
time-to-event traits as exposures in MR analyses.

Researchers have also drawn on other generalized linear models for analysis in MR 
studies. However, this area is relatively unexplored. @hazewinkel_mendelian_2022 treated hospitalizations as a count variable and 
modeled it with quasi-Poisson methods. @allman_mendelian_2022 characterized statistical properties, including bias and 
power, with simulations of count random variables and binary random variables. With growing interest
in modeling counts in biomedical data, where molecular phenotyping technologies now acquire RNA molecule counts and protein abundances, 
there may be opportunities for methods innovations in a generalized linear models framework for MR studies.

See also @tchetgen_tchetgen_genius_2021

>>>>

Genome-wide association studies (GWAS) present unique opportunities for 
discovery of causal relationships via MR. A GWAS
interrogates millions of single nucleotide polymorphisms (SNPs) to infer which affect the trait [@the_wellcome_trust_case_control_consortium_genome-wide_2007;@klein_complement_2005].
In the nearly 17 years since publication of early studies, researchers have reported 
thousands of novel SNP-trait associations from GWAS [@klein_complement_2005;@welter_nhgri_2014;@visscher_10_2017;@visscher_five_2012].

@@

Recently, multiple teams of investigators have analyzed time-to-event traits in MR studies [@burgess_commentary_2015;@arvanitis_linear_2021;@tikkanen_body_2019;@he_causal_2018;@guo_body_2017]. 
@guo_body_2017, for example, studied post-diagnosis survival time in women with breast cancer. They 
treated survival as a time-to-event outcome, which enabled them to model it with a proportional 
hazards model [@cox_regression_1972] in two-stage MR with individual-level data. Similarly, 
@he_causal_2018, in a study of cardiovascular disease, applied proportional hazards regression when 
examining time to stroke as an outcome variable. In very recent work, @tchetgen_tchetgen_genius_2021
reported methods, MR GENIUS, that are robust to violations of the exclusion restriction assumption (under 
many assumed MR data generating processes). They considered binary, continuous, and time-to-event 
outcomes within their framework. We are unaware of MR studies that treat 
the exposure variable as a time-to-event, but it's conceivable that future research will consider 
time-to-event traits as exposures in MR analyses.

Researchers have also drawn on other generalized linear models for analysis in MR 
studies. However, this area is relatively unexplored. @hazewinkel_mendelian_2022 treated hospitalizations as a count variable and 
modeled it with quasi-Poisson methods. @allman_mendelian_2022 characterized statistical properties, including bias and 
power, with simulations of count random variables and binary random variables. With growing interest
in modeling counts in biomedical data, where molecular phenotyping technologies now acquire RNA molecule counts and protein abundances, 
there may be opportunities for methods innovations in a generalized linear models framework for MR studies [@hwang_single-cell_2018;@ozsolak_rna_2011;@domon_mass_2006;@bantscheff_quantitative_2007].
